Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-02-24 paper.tex
Improved treatment of representations

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
The fundamental weights $P=\{\varpi_1,\ldots\varpi_\ell\}$ are defined as the basis dual to the simple coroots, i.\,e. $\langle \varpi_i, \alpha_j^{\vee} \rangle = \delta_{ij}$.

@@

In the sequel we use elementary facts about representations of Chevalley groups (see e.\,g.~\cite{PSV98,VP} for a more detailed introduction).
Let $L$ be the complex simple Lie algebra of type $\Phi$.
To each element $\mu$ of the cone of dominants weights $\mathbb{Z}_{\geq 0}\langle P\rangle$
 one can associate an irreducible finite dimensional complex representation $\pi \colon L \to \mathrm{gl}(V)$ with highest weight $\mu$.
For our purposes it suffices to restrict our attention only to fundamental representations of $L$ (i.\,e. representations with highest weight $\varpi_i$),
 which, moreover, we can assume to be {\it basic} in the sense of~\cite[\S~I.2]{Ma69}.

Denote by $\Lambda(\pi)$ the set of weights of $\pi$. Recall that $V$ admits weight decomposition $V = V_0\oplus \bigoplus_{0\neq\lambda\in \Lambda(\pi)} V_\lambda$,
 moreover, thanks to our assumptions, all $V_\lambda$'s corresponding to nonzero $\lambda$ are $1$-dimensional.
By Chevalley--Ree theorem in any finite-dimensional representation $(\pi, V)$ of $L$ one can choose a basis $v_\lambda \in V_\lambda$,
 with the property that $\pi(e_\alpha^{(m)})(v_\lambda)$ is an integral combination of $v_\lambda$ for every $m\in \mathbb{Z}_{\geq 0}$, $\alpha\in \Phi$.

Consequently one can construct a representation $\G_{sc}(\Phi, R)$ on the free $R$-module $V_\mathbb{Z}\otimes R$ called a {\it Weyl module}.
For us it will only be relevant that in the case of a basic representation, there are particularly nice formulae describing the action
 of elementary root unipotents $x_\alpha(\xi)$ on weight spaces $V_\lambda$, $\lambda \in \Lambda(\pi)$, see~\cite[Lemme~2.3]{Ma69},
 which can be, moreover, visualized using the technique of {\it weight diagrams} from \cite{PSV98}.

The following lemma is, in essence, a relative version of Chevalley--Matsumoto decomposition which can be easily deduced from the absolute version (cf. \cite[Theorem~1.3]{St78}).
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be a basic fundamental representation of $\G_{sc}(\Phi, R)$ with highest weight $\varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\varpi_{s}}=1$, then $g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I)$. \end{lemma}
Here by $v^+$ we denote the highest vector of $\pi$ and by $v_\varpi$ the coordinate of $ $

@@

In this section we work with standard vector representations of classical groups (i.\,e. representations with highest weight $\varpi_1$).
It will be convenient for us to number the weights of these representations as in~\cite[\S~1B]{St78}:
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ such that $\lambda_1-\lambda_2\in \Phi$.
It will be convenient for us to write $x_{\lambda_1, \lambda_2}(\xi)$ instead of $x_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ we have $x_{1, 2}(\xi)=x_{1-2}(\xi)=x_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = x_{\alpha_1}(\xi)$.

@@

The following lemma is the key point of the proof where stability assumptions are invoked.
\begin{lemma}\label{lemma:Stein_reduction}
Under the assumptions of \cref{thm:DennisVaserstein} one has $X_{-\alpha_r}(R) \subseteq N(A_{rs}).$
\end{lemma}
\begin{proof}
Let $\Phi$, $r$ and $s$ be as in Table~\ref{table:dv-reps} and let $\pi \colon \G(\Phi, R) \to \mathrm{GL}(V_R)$ be the Weyl module with highest weight $\varpi_s$.
The type and the dimension of $\pi$ are listed in Table~\ref{table:dv-reps}. %, see also~\cite{PSV98} where the weight diagrams of these representations are depicted.

@@

Now let us show that $A = A_{rs}$.
Let $g$ be an element of $\E(\Delta_r, R, I)$. Applying \cref{lemma:uraction} to the subsystem $\Delta_r$, we find
$x\in\U(\Delta_r^+, I)$ and $y\in \U(\Delta_r^-, I)$ such that $yx\cdot g \in B$, therefore
\begin{equation*} \U(\Sigma^+_r, I) \cdot \U(\Sigma^-_r, I) \cdot g = \U(\Sigma^+_r, I) x^{-1} \cdot \U(\Sigma^-_r, I)^{x^{-1}} y^{-1} (yxg) \subseteq \U(\Phi^+, I) \cdot \U(\Phi^-, I) \cdot B. \end{equation*}

@@

For the rest of this subsection $k$ denotes a global field and $S$ some finite set of places on $k$. 
Let $\mathcal{O}_S$ be the Dedekind ring of arithmetic type defined by $S$ and let $I$ be an ideal of $\mathcal{O}_S$.

\begin{lemma}\label{lemma:width-dedekind}
Let $\Phi$ be an irreducible classical root system of rank $\ell \geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi, \mathcal{O}_S, I)$ has finite width with respect to the generating set $\mathcal{Z}(\Sigma_\ell)$.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leqslant \asr(\mathcal{O}_S) \leqslant 2$. 
By \cref{cor:bass-kolster-iterated} we can write any element 
of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators from $\mathcal{Z}(\Sigma_\ell)$ and one element of 
$G_0 = \G(\{\alpha_\ell, -\alpha_\ell\}, \mathcal{O}_S, I)\cong\SL(2, \mathcal{O}_S, I)$.
Consequently, to prove the statement of the lemma it suffices to express every element 
$g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$
as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

@@

By \cref{lemma:Stepanov-ideal} the group $\E\left(\Phi, \mathcal{O}_S, I^2\right)$ is contained in $\E(\Phi, I)$ for any root system $\Phi\neq\rC_\ell$ of rank $\geqslant2$.
Notice that under the assumptions of the lemma it is known that $\E(\Phi, I)$ has finite width with respect to $\mathcal{X}$, see~\cite[Theorem~3.3]{TavgenThesis}.

>>>>

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\beta, \beta)$.
Recall that fundamental weights $\varpi_1,\ldots,\varpi_\ell$ are characterized by the property that $\langle \varpi_i, \alpha_j \rangle = \delta_{ij}$.

@@

In the sequel we use elementary facts about representations of Chevalley groups (see e.\,g.~\cite{PSV98,VP} for a more detailed introduction).
Recall that, by definition, the {\it cone of dominant weights} $P_{++}(\Phi)$ consists of all nonnegative integral linear combinations of fundamental weights $\varpi_1,\ldots \varpi_\ell$.
To each element $\mu \in P_{++}(\Phi)$ one can associate a representation $\pi$ of $\G(\Phi, R)$ on a free $R$-module $V = V(\mu)$, called {\it Weyl module}.
For our purposes it suffices to restrict attention to representations with fundamental highest weight (i.\,e. $\mu=\varpi_i$ for some $i$), which are, moreover, {\it basic} in the sense of~\cite[\S~I.2]{Ma69}.
The latter condition guarantees, in particular, that all weight subspaces $V^\lambda \leq V$ corresponding to nonzero weights $\lambda$ have dimension $1$.
Denote by $\Lambda(\pi)$ the set of weights of $\pi$ with multiplicities (i.\,e. the zero weight is repeated $\dim(V^0)$ times).

By the very construction of $V(\mu)$, there is a basis $\{e_\lambda\}$, $e_\lambda\in V^\lambda$ parameterized by elements $\lambda\in\Lambda(\pi)$ called {\it admissible basis}.
The key property of basic representations which we use in the sequel is that there are particularly simple formulae describing the action
 of elementary root unipotents $x_\alpha(\xi)$ on the vectors of admissible basis (see~\cite[Lemme~2.3]{Ma69}),
 which, moreover, can be visualized using the technique of {\it weight diagrams} (see e.\,g.~\cite{PSV98}).

We denote by $v^+$ the {\it highest weight vector}, i.\,e. the basis vector $e_\mu$ corresponding to the highest weight $\mu$ of $V(\mu)$.
For $v\in V(\mu)$ we denote by $v_\lambda$ the coordinate of $v$ in the expansion of $v$ in $\{e_\lambda\}$, moreover if $\pi$ is a faithful representation we can 
identify $g\in \G(\Phi, R)$ with the matrix $(g_{\lambda, \nu})$, where $g_{\lambda, \nu} = (\pi(g)\cdot e_\nu)_{\lambda}$.
The following lemma is, in essence, a relative version of Chevalley--Matsumoto decomposition which can be easily deduced from the absolute statement (cf. \cite[Theorem~1.3]{St78}).
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be a basic fundamental representation of $\G_{sc}(\Phi, R)$ with highest weight $\mu = \varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\mu}=1$, then $g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I)$. \end{lemma}

@@

In this section we work with natural representations of classical groups (i.\,e. representations with highest weight $\varpi_1$).
It will be convenient for us to number the weights of these representations as in~\cite[\S~1B]{St78}:
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ etc.

Let $\lambda_1, \lambda_2 \in \Lambda(\pi)$ be a pair of weights of a representation $\pi$ such that $\lambda_1-\lambda_2\in \Phi$.
For notational convenience we write $x_{\lambda_1, \lambda_2}(\xi)$ instead of $x_{\lambda_1-\lambda_2}(\xi)$.
For example, for $\Phi=\rA_\ell$ the notation $x_{1, 2}(\xi)$ has the same meaning as $x_{\varpi_1 - \varpi_1 + \alpha_1}(\xi) = x_{\alpha_1}(\xi)$.

@@

The following lemma is the key point of the proof where stability assumptions are invoked.
\begin{lemma}\label{lemma:Stein_reduction}
Under the assumptions of \cref{thm:DennisVaserstein} one has $X_{-\alpha_r}(R) \subseteq N(A_{rs}).$
\end{lemma}
\begin{proof}
Let $\Phi$, $r$ and $s$ be as in Table~\ref{table:dv-reps} and let $\pi$ be the representation of $\G(\Phi, R)$ on the Weyl module $V(\mu)$ with highest weight $\mu=\varpi_s$.
The type and the dimension of $\pi$ are listed in Table~\ref{table:dv-reps}. %, see also~\cite{PSV98} where the weight diagrams of these representations are depicted.

@@

We claim that $A = A_{rs}$. Indeed, let $g$ be an element of $\E(\Delta_r, R, I)$. 
Applying \cref{lemma:uraction} to the subsystem $\Delta_r$, we find $x\in\U(\Delta_r^+, I)$ and $y\in \U(\Delta_r^-, I)$ such that $yx\cdot g \in B$.
Consequently we have the following inclusion:
\begin{equation*} \U(\Sigma^+_r, I) \cdot \U(\Sigma^-_r, I) \cdot g = \U(\Sigma^+_r, I) x^{-1} \cdot \U(\Sigma^-_r, I)^{x^{-1}} y^{-1} (yxg) \subseteq \U(\Phi^+, I) \cdot \U(\Phi^-, I) \cdot B. \end{equation*}

@@

For the rest of this subsection $k$ denotes a global field and $S$ some finite set of places on $k$. 
Let $\mathcal{O}_S$ be the Dedekind ring of arithmetic type defined by $S$ and let $I$ be an ideal of $\mathcal{O}_S$.
\begin{lemma}\label{lemma:width-dedekind}
Let $\Phi$ be an irreducible classical root system of rank $\ell \geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi, \mathcal{O}_S, I)$ has finite width with respect to the generating set $\mathcal{Z}(\Sigma_\ell)$.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leqslant \asr(\mathcal{O}_S) \leqslant 2$. 
By \cref{cor:bass-kolster-iterated} we can write any element 
of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators from $\mathcal{Z}(\Sigma_\ell)$ and one element of 
$G_0 = \G(\{\alpha_\ell, -\alpha_\ell\}, \mathcal{O}_S, I)\cong\SL(2, \mathcal{O}_S, I)$.
Consequently, to prove the statement of the lemma it suffices to express every element 
$g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$
as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

@@

By \cref{lemma:Stepanov-ideal} the group $\E\left(\Phi, \mathcal{O}_S, I^2\right)$ is contained in $\E(\Phi, I)$ for any root system $\Phi\neq\rC_\ell$ of rank $\geqslant2$.
Notice that under the assumptions of the lemma a deep result of O.~Tavgen asserts that $\E(\Phi, I)$ has finite width with respect to $\mathcal{X}$, see~\cite[Theorem~3.3]{TavgenThesis}.
